Reproducible Research: Peer Assessment 1
========================================================


## Loading and preprocessing the data


  #1. Load the data

#read in "activity.csv"

```{r, echo = TRUE}
setwd("C:/Users/ACLINN01/Documents")
activity <- read.csv("activity.csv")
```

  # 2. Process/transform the data into a format suitable for your analysis

#change the column "date" to native R date format

```{r, echo = TRUE}
activity$date<-as.Date(activity$date,"%Y-%m-%d")

library(data.table)
library(ggplot2)
```
## What is mean total number of steps taken per day?


  #1. Make a histogram of the total number of steps taken each day

#summarize steps by date, omitting NA values

```{r, echo = TRUE}
activity_daily<-data.table(na.omit(activity))[,list(total_steps=sum(steps)), by=(date)]
```

#open a PNG and plot the values

```{r ggplot, echo = TRUE, results='hide', message=FALSE, warning=FALSE}
ggplot(activity_daily,aes(x=date,y=total_steps))+geom_bar(stat='identity') +
ggtitle("Total Steps by Day") + ylab("Steps")  + xlab("Date")
```

  #Calculate and report the mean and median total number of steps taken per day

#calculate the daily mean and median values

```{r, echo = TRUE}
mean(activity_daily$total_steps, na.rm = TRUE)
median(activity_daily$total_steps, na.rm = TRUE)
```


## What is the average daily activity pattern?

  #1. Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

#average steps by interval, omitting NA values

```{r, echo = TRUE}
activity_int<-data.table(na.omit(activity))[,list(avg_steps=mean(steps)), by=(interval)]
```

#open a PNG and plot the values

```{r}
ggplot(activity_int, aes(interval, avg_steps)) + geom_line() +
  ggtitle("Average Steps per Interval") + ylab("Steps")  + xlab("Inverval")
```


  #2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

#subset the data to return the inteval and avg. steps for the interval with the most avg steps

```{r, echo = TRUE}
activity_int[avg_steps==max(activity_int$avg_steps)]
```


## Imputing missing values


  #1. Calculate and report the total number of missing values in the dataset 

```{r, echo = TRUE}
sum(is.na(activity$steps))
```

  #2. Devise a strategy for filling in all of the missing values in the dataset. 

#subset only the missing values and non-missing values seperately

```{r, echo = TRUE}
mis<-subset(activity,is.na(activity$steps))
nomis <- subset(activity,!is.na(activity$steps))
```

#merge the missing values data table with the interval averages
```{r, echo = TRUE}
mis2 <- data.table(subset(merge(mis, activity_int, by="interval"),select = -steps))
```

#rename and reorder columns

```{r, echo = TRUE}
setnames(mis2,"avg_steps","steps")
setcolorder(mis2,c("steps","date","interval"))
```

  #3. Create a new dataset that is equal to the original dataset but with the missing data filled in.

#union the mis data set with replaced values with the nomis data

```{r, echo = TRUE}
finaldat <- rbind(mis2, nomis)
```

  #4. Make a histogram of the total number of steps taken each day and Calculate and report the mean 
  #   and median total number of steps taken per day. 

#summarize steps by date

```{r, echo = TRUE}
activity_daily2<-data.table(finaldat)[,list(total_steps=sum(steps)), by=(date)]
```

#open a PNG and plot the values

```{r echo = TRUE, results='hide', message=FALSE, warning=FALSE}
ggplot(activity_daily2,aes(x=date,y=total_steps))+geom_bar(stat='identity') +
ggtitle("Total Steps by Day") + ylab("Steps")  + xlab("Date")
```

  #a. Do these values differ from the estimates from the first part of the assignment?


mean: NA values removed
```{r, echo = TRUE}
mean(activity_daily$total_steps, na.rm = TRUE)
```
mean: NA values replaced
```{r, echo = TRUE}
mean(activity_daily2$total_steps, na.rm = TRUE)
```
median: NA values removed
```{r, echo = TRUE}
median(activity_daily$total_steps, na.rm = TRUE)
```
median: NA values replaced
```{r, echo = TRUE}
median(activity_daily2$total_steps, na.rm = TRUE)
```


  #b. What is the impact of imputing missing data on the estimates of the total daily number of steps?


mean values were used to impute missing values, so means do not change. 
however, additional values are present so median values are affected


## Are there differences in activity patterns between weekdays and weekends?

  #1. Create a new factor variable in the dataset with two levels - 
  #   "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

#add weekday

```{r, echo = TRUE}
finaldat$wd <- weekdays(finaldat$date)
```

#replace values with "Weekend" and "Weekday"

```{r, echo = TRUE}
finaldat$wd[finaldat$wd %in% c("Saturday","Sunday")]<-"Weekend"
finaldat$wd[finaldat$wd %in% c("Monday","Tuesday","Wednesday","Thursday","Friday")]<-"Weekday"
```
  #2. Make a panel plot containing a time series plot of the 5-minute interval and the average number 
  #   of steps taken, averaged across all weekday days or weekend days.

#average steps by interval, subsetted by weekend or weekday

```{r, echo = TRUE}
activity_int3<-data.table(subset(finaldat, wd == "Weekend"))[,list(avg_steps=mean(steps)), by=(interval)]
activity_int4<-data.table(subset(finaldat, wd == "Weekday"))[,list(avg_steps=mean(steps)), by=(interval)]

p1<- 
ggplot(activity_int3, aes(interval, avg_steps)) + geom_line() +
ggtitle("Average Steps per Interval - Weekend") + ylab("Steps")  + xlab("Inverval")

p2<- 
ggplot(activity_int4, aes(interval, avg_steps)) + geom_line() +
ggtitle("Average Steps per Interval - Weekday") + ylab("Steps")  + xlab("Inverval")


#open a PNG, plot two charts


library(grid)

grid.newpage()
pushViewport(viewport(layout = grid.layout(2, 1)))
print(p1, vp = viewport(layout.pos.row = 1, layout.pos.col = 1))
print(p2, vp = viewport(layout.pos.row = 2, layout.pos.col = 1))
```
